---
title: "week7_transform"
author: "Pushkar"
date: "January 22, 2019"
output: html_document
---

### Needed packages
```{r,message=FALSE}
library(nycflights13)
library(tidyverse)
#dplyr package is included in tidyverse package
```

This week we will be working on some basic functions in the dplyr package.
## filter() - used to select rows/ observations
```{r}
#get all flights on February 2
filter(flights, month == 2, day == 2)
#get all flights that departed in November or December
filter(flights, month == 11 | month == 12)
#find flights that weren't delayed (on arrival or departure) by more than two hours
filter(flights, !(arr_delay > 120 | dep_delay > 120))
#find flights that flew to Houston (IAH or HOU)
filter(flights, dest == 'IAH'| dest == 'HOU'  )
```
## arrange() - used to rearrange the rows/ observations by columns specified
```{r}
arrange(flights, year, month, day)

#arrange flights in decreasing order of departure delay
arrange(flights, desc(dep_delay))

```
## select() - used to select columns
```{r}
# Select columns year, month, day, dep_delay, arr_time
select(flights, year, month, day, dep_delay, arr_time)

# Select all columns between year and dep_delay
select(flights, year:dep_delay)

# Select all columns except those from year, month,  day , dep_time
select(flights, -(year:dep_time))


```
## mutate() - create new variables by manipulating existing functions 
```{r}
#use mutate() to find gain in delay and speed of flight
gs_flights <- mutate(flights,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)

gs_flights
```
## summarise() - gives a single row summary, typically of a grouped data
```{r}
# find mean departure delay, ignore na values
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))

```
## Combining multiple operations with the pipe
```{r}
#find mean dep_delay per day while ignoring na values
flights %>% group_by(year, month, day) %>% mutate(delay = mean(dep_delay, na.rm = TRUE))

# Find popular destinations which have greater than 200 flights:
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 200)
popular_dests

```
