---
title: "week6: Pirates and Superheroes"
author: "Pushkar"
date: "January 17, 2019"
output: html_document
---


##1. Combine the data into a single dataframe.
```{r, echo= TRUE, include=TRUE}
Name = c("Astrid","Lea","Sarina","Remon","Letizia","Babice","Jonas","Wendy","Niveditha","Gioia") 

Sex = c("F","F","F","M","F","F","M","F","F","F")

Age = c(30,25,25,29,22,22,35,19,32,21)

Superhero = c("Batman","Superman","Batman","Spiderman","Batman","Antman","Batman","Superman","Maggott","Superman")

Tattoos = c(11,15,12,5,65,3,9,13,900,0)

pirates = data.frame(Name, Sex, Age, Superhero, Tattoos)

```

```{r, echo= FALSE, include= FALSE  }
str(pirates)
```

## 2. What is the median age of the 10 pirates?
```{r, echo= TRUE, include=TRUE }
median(pirates$Age)
```

## 3. What was the mean age of female and male pirates separately?
```{r}
'meanFemale'
mean(pirates$Age[pirates$Sex == "F"])
'meanMale'
mean(pirates$Age[pirates$Sex == "M"])

```
## 4. What was the most number of tattoos owned by a male pirate?
```{r}
max(pirates$Tattoos[pirates$Sex == "M"])
```

## 5. What percent of pirates under the age of 32 were female?
```{r}
plBelow <- pirates[pirates$Age < 32, ]
pFemale <- plBelow[plBelow$Sex == "F",]

100 * nrow(pFemale)/ nrow(plBelow)
```

## 6. What percent of female pirates are under the age of 32?
```{r}
pFem <- pirates[pirates$Sex == "F",]
pBelow <- pFem[pFem$Age < 32, ]

100 * nrow(pBelow) / nrow(pFem)
```

## 7. Add a new column tattoos.per.year showing how many tattoos each pirate has for each year in their life.
```{r}
pirates$Tattoos <- as.numeric(pirates$Tattoos)
pirates$TattoosPerYear <- pirates$Tattoos / pirates$Age

```
## 8. Which pirate had the most number of tattoos per year?
```{r}
pirates$Name[pirates$TattoosPerYear == max(pirates$TattoosPerYear)]
```
## 9. What are the names of the female pirates whose favorite superhero is Superman?
```{r}
pirates$Name[pirates$Superhero == "Superman"]
```
## 10. What was the median number of tattoos of pirates with age > 20 whose favorite superhero is Spiderman?
```{r}
pAbove <- pirates[pirates$Age > 20,]
paSuperhero <- pAbove[pAbove$Superhero == "Spiderman",]
median(paSuperhero$Tattoos)
```


