---
title: "week9_plotting"
author: "Pushkar"
date: "January 27, 2019"
output: html_document
---
```{r, message=FALSE}
library(ggplot2)
library(tidyverse)
```

##Data Visualization
This week we will be working on using the ggplot() function to plot data
We will use the inbuilt diamonds dataset 
```{r}
#Use bar chart to see distribution of a categorical variable
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```
```{r}
#Use histogram to see distribution of a continuos variable
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```
```{r}
#Use scatterplot to see relationship between carat and price of diamonds
plot(diamonds$carat, diamonds$price)

```
Between 0 to 1, it appears like increase in carat leads to pricier diamonds

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) + ggtitle("weird diamonds$y plot")
```
There's something weird with the plot. The y axis is taking a lot of space even if we do not see any points on the screen. There might be some outliers which are not clearly visible here.To make it easy to see the unusual values, we need to zoom to small values of the y-axis with coord_cartesian()
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 60))
```
This allows us to zoom in so that we can see outliers easily.

There are two ways to remove unususal values. One is to drop them and the other option, the better one is to 
use mutate function.
```{r}
diamonds2 <- diamonds %>%   mutate(y = ifelse(y < 3 | y > 20, NA, y))

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()+ ggtitle("Diamonds$x vs Diamonds$y")
```
geom_point() will always show if values have been removed. na.rm function can be used to suppres that.
```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)
```
let's explore how the price of a diamond varies with its quality. We will need to explore the distribution of a continuous variable broken down by a categorical variable.
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 600)
```
It's hard to see the difference in distribution because the overall counts differ so much:

```{r}
ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

Lets flip on the y-axis to show density.
```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```
it appears that fair diamonds (the lowest quality) have the highest average price!

Let's take a look at the distribution of price by cut using geom_boxplot():

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()

```
It looks like better quality diamonds are cheaper on average. 

To visualise the covariation between categorical variables, you'll need to count the number of observations for each combination. geom_count() exists for just that.
```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
```
The size of each circle in the plot displays how many observations occurred at each combination of values.

