---
title: "week6_functions"
author: "Pushkar"
date: "January 20, 2019"
output: html_document
---
### example data
```{r}
x <- c('a', 'b', 'c', 'd')
y <- c('e', 'f', 'g', 'h')
z <- c('i', 'j', 'k', 'l')
```

### cbind() - used to combine vector or data frame by columns
```{r}
cbind(x,y,z)
```

### rbind() - used to combine vector or data frame by rows
```{r}
rbind(x,y,z)
```
### matrix(x, nrow, ncol, byrow) - create a matrix of order nrow X ncol
byrow option is used to decide if the elements are filled row-wise (byrow = 0) or column-wise (1)
```{r}
matrix( x <- 22:29, nrow = 2 , ncol = 4, byrow = 1)
```
### data.frame() - create a data frame 
A data frame is used for storing data tables. It is a list of vectors of equal length.
```{r}
n = c(2, 3, 5) 
s = c("aa", "bb", "cc")
b = c(TRUE, FALSE, TRUE)
data.frame(n,s,b)
```
#### stringsAsFactors = FALSE
By default, when you create a data frame, string vectors are interpreted as factors
To avoid that we mention "stringsAsFactors = FALSE" in the argument of the data frame
```{r}
survey <- data.frame("index" = c(1, 2, 3, 4, 5),
                     "sex" = c("m", "m", "m", "f", "f"),
                     "age" = c(99, 46, 23, 54, 23))
str(survey)
```
We can see that when not mentioned, sex vector is interpreted as a factor.
Let us remedy this.
```{r}
survey <- data.frame("index" = c(1, 2, 3, 4, 5),
                     "sex" = c("m", "m", "m", "f", "f"),
                     "age" = c(99, 46, 23, 54, 23),
                     stringsAsFactors = FALSE
                     )
str(survey)
```

### head() - view first six rows of a dataset
```{r}
head(ToothGrowth)
```
### tail() - view las six rows of a dataset
```{r}
tail(ToothGrowth)
```
### View() - opens all of the dataset in a new window
```{r}
View(ToothGrowth)
```
### summary() - gives stats like min, 1st quart., median, mean, 3rd quart. and max for all cols in the dataset
```{r}
summary(ToothGrowth)
```
### str() - gives information on the datatypes of columns in a dataset and also shows first few elements
```{r}
str(ToothGrowth)
```
### names() - outputs a string vector of all the column names in a dataset
```{r}
names(ToothGrowth)
```
### Changing column names
```{r}
#names(df)[1] <- "number"
#commented out cause running it multiple time throws error 
#because the same name is already assigned once
```

It's possible to mess up naming this way if we have too many columns
We can play safe by using logical indexing.
Here's how we do it.
```{r}
#names(df)[names(df) == "s"] <- "strings"
#commented out cause running it multiple time throws error 
#because the same name is already assigned once

```

### Slicing with [, ]
#### lets try slicing with PlantGrowth dataset
```{r}
# lets have a look at the dataset
str(PlantGrowth)
head(PlantGrowth)

# Return  the rows 1-26 and column 2 
PlantGrowth[1:26,2]

# Return first row across all cols
PlantGrowth[1,]

# Return first col across all rows
PlantGrowth[,1]

# Return a dataframe with only rows where group = 'trt1'
PlantGrowth[PlantGrowth$group == 'trt1',]

# Return a dataframe with weight < 5 and group = 'trt2'
PlantGrowth[PlantGrowth$weight < 5 & PlantGrowth$group == 'trt2', ]

```
### Slicing with subset(x, subset = , select = )
x = dataframe,
subset = logical vector which tells us which rows to keep,
select = tells us cols to keep
```{r}
# Return weight of the plant where weight < 5 and group = 'trt2'
subset(x = PlantGrowth, subset = weight < 5 & group == 'trt2', select = 'weight')

```

### with() - time saver when we need to use multiple cols of a data frame
```{r}
health <- data.frame("age" = c(32, 24, 43, 19, 43),
                     "height" = c(1.75, 1.65, 1.50, 1.92, 1.80),
                     "weight" = c(70, 65, 62, 79, 85))

#calculate ratio of height to weight
with(health, height/weight)
```
